\\textsc{Mathematical Description}

### 2023-06-07 100-math-description/GR-sslcm-math-description.Rmd
Create/implement shortcut commands
e.g., \NOR instead of having to type \mathrm{NOR}

---
title: "\\textsc{Mathematical Description}"
subtitle: "\\textsc{State-Space Life Cycle Model for Grande Ronde Spring Chinook Salmon}"
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
  # bookdown::word_document2
fontfamily: mathpazo
geometry: [top=1in,bottom=1in,left=0.25in,right=1.75in]
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: cites.bib
link-citations: true
biblio-style: apalike
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \definecolor{my-blue}{HTML}{E6F0FF}
  - \usepackage[textwidth=1.5in]{todonotes}
  - \setuptodonotes{backgroundcolor=my-blue,textcolor=black,linecolor=ProcessBlue,bordercolor=my-blue,size=footnotesize}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
  - \renewcommand{\eqref}[1]{\ref{#1}}
  - \renewcommand*{\log}{\mathrm{log}\hspace{-0.20em}}
  - \renewcommand*{\max}{\mathrm{max}\hspace{-0.20em}}
  - \newcommand{\logit}{\mathrm{logit}\hspace{-0.20em}}
  - \newcommand{\ilogit}{\mathrm{logit}^{-1}\hspace{-0.20em}}
  - \newcommand{\MVNdist}{\mathcal{MVN}\hspace{-0.20em}}
  - \newcommand{\Ndist}{\mathcal{N}\hspace{-0.20em}}
  - \newcommand{\Bdist}{\mathcal{B}\hspace{-0.20em}}
  - \newcommand{\Mdist}{\mathcal{M}\hspace{-0.20em}}
  # - \newcommand{\fall}{\mathrm{fall}}
  # - \newcommand{\spring}{\mathrm{spring}}
  # - \newcommand{\NOR}{\mathrm{NOR}}
  # - \newcommand{\HOR}{\mathrm{HOR}}
---

@@

To facilitate later analyses (not documented here) investigating the effects of habitat restoration and climate change on population dynamics, we modeled parr capacity as a function of weighted usable habitat length ($\mathrm{WUL}_j$)^[Derivation of $\mathrm{WUL}_j$ is too detailed to describe fully here.
Briefly, the relative ability of stream segments with different habitat characteristics to hold rearing parr was expressed in terms of the density of fish predicted to reside there by a complex generalized linear mixed model.
The model was fitted to density estimates based on detectability-adjusted [@staton-etal-2022] summer snorkel counts with local habitat characteristics (e.g., pool frequency, water temperature, large wood density, etc.) as predictor variables.
The weight of each stream segment was assigned based on the predicted density relative value to the segment with the highest value; $\mathrm{WUL}_j$ was then equal to the sum of segment-specific weight and length within the known spawning and rearing extent for each population.].
We modeled parr recruitment capacity ($\beta_j$) as a function of this $\mathrm{WUL}_j$ metric:

\begin{equation}
\log\left(\beta_j\right) \sim \Ndist\left[\log\left(\lambda \cdot \mathrm{WUL}_j\right), \sigma_{\beta} \right]
(\#eq:beta-random)
\end{equation}

where $\lambda$ is the expected change in parr capacity per 1 km change in weighted usable habitat length, and $\sigma_{\beta}$ is the log-normal standard deviation of variability in this relationship not captured by $\mathrm{WUL}$ values.

@@

\begin{equation}
\log\left(\dot{L}^{P^{b}}_{y,j}\right) = \omega_{0,j} + \omega_{1,j} \cdot \log\left(\frac{E_{y,j}}{\mathrm{WUL}_j}\right)
(\#eq:get-L-Pb-dot)
\end{equation}

@@

Like many populations of stream-type spring Chinook salmon in the Columbia River basin \todo{Insert references about other CRB populations displaying the fall/spring migratory strategy}, there is life history diversity among the juveniles with respect to their migratory phenology and habitat types occupied for rearing overwinter.
Some portion of the summer parr ($P^{b}_{y,j}$)  migrate out of the headwaters near where they were spawned in the fall and rear farther downstream (termed "fall migrants" and indexed by $i = \mathrm{fall}$) and the remaining portion migrate out the following spring (termed "spring migrants" and indexed by $i = \mathrm{spring}$).
Due to this difference in migratory phenology, these two groups of fish are monitored separately \todo{Insert a reference to a recent ODFW screw trap monitoring report where it is stated (or implied) that migratory strategies are monitored separately.} depending on when they pass the rotary screw trap located in each tributary\todo{Create a map and include reference it here}.
To enable building an observation model that could accommodate this feature of the populations in our model, we apportioned summer parr to each of these two strategies:

@@

where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
We assumed the expected value of this proportion for fall migrants ($\dot{\pi}_{i = \mathrm{fall},j}$) was time-constant and modeled the realized values as having a multivariate logit-normal distribution with covariance matrix $\Sigma_{\pi_{i = \mathrm{fall}}}$:

\begin{equation}
\logit\left(\boldsymbol{\pi}_{y,i = \mathrm{fall},1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\pi}}_{i = \mathrm{fall},1:n_j}\right), \Sigma_{\pi_{i = \mathrm{fall}}}\right]
(\#eq:pi-random)
\end{equation}

Since there are only two migratory strategies, we obtained the proportion that are spring migrants as the complement ($\pi_{y,i = \mathrm{spring},j} = 1 - \pi_{y,i = \mathrm{fall},j}$).

@@

Since hatchery-origin smolts are released into the population in spring (see Section \@ref(hatchery-inputs)), we introduce the origin index ($o$) starting at the in-basin smolt stage prior to out-migration ($M^{b}_{y,i,o,j}$).
The abundance of natural-origin (indexed by $o=\mathrm{NOR}$) smolt was then:

\begin{equation}
M^{b}_{y,i,o=\mathrm{NOR},j} = P^{a}_{y,i,j} \cdot \phi^{P^{a} \to M^{b}}_{y,i,j}
(\#eq:get-Mb-NOR)
\end{equation}

$\rho_{\pi_{i=\mathrm{fall},j,j^{\prime}}}$

@@

For populations with hatchery supplementation (CAT, LOS, and UGR), hatchery-origin smolt releases (assumed known without error \todo{Insert reference to a document that gives info about these smolt releases}) were stored in the variable $M^{b}_{y,i=\mathrm{spring},o=\mathrm{HOR},j}$; we assumed no hatchery-origin parr were released the prior year, thus $M^{b}_{y,i=\mathrm{fall},o=\mathrm{HOR},j} = 0$.
For all subsequent life stages, we tracked the abundance (and often survival) of hatchery-origin fish separately from natural-origin fish.

@@

All in-basin smolt ($M^{b}_{y,i,o,j}$) migrate to the ocean in the spring of the year they become age-2 regardless of their migratory strategy ($i$).
We divide mortality experienced during this migration into two stages: (_i_) from in-basin to Lower Granite Dam (LGR), the first in a series of eight dams\todo{Insert reference to a document describing dams} encountered in the downstream migration, and (_ii_) from LGR to the ocean.
For parameter identifiability in mortality stage (_i_), we were required to assume that survival from in-basin to LGR was identical between natural-origin fall and spring migrants (i.e., $\phi^{M^{b} \to M^{a}}_{y,i=\mathrm{fall},o=\mathrm{NOR},j} = \phi^{M^{b} \to M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}$, we thus omit the $i$ dimension when defining these quantities below), but we were able to model natural-origin survival separately from hatchery-origin because they are monitored separately (see Section \@ref(survival-data-sources)).

@@

\begin{equation}
\logit\left(\dot{\phi}^{M^{b} \to M^{a}}_{y,o=\mathrm{NOR},j}\right) = \tau_{0,j} + \tau_{1,j} \cdot L^{*M^{b}}_{y,j}
(\#eq:get-phi-Mb-Ma-dot)
\end{equation}

where $L^{*M^{b}}_{y,j}$ has been scaled and centered by subtracting the historical average observed smolt mean length for population $j$ from $L^{M^{b}}_{y,j}$ and dividing by the historical observed standard deviation (just as for $L^{*P^{b}}_{y,j}$ for standardized parr mean length).
For hatchery-origin smolt, we assumed this expected out-of-basin survival term was time-constant (i.e., all $\dot{\phi}^{M^{b} \to M^{a}}_{y,o=\mathrm{HOR},j}$ were equal within a population).
For each origin separately, we modeled realized out-of-basin migration survival as a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{M^{b} \to M^{a}}}$ (assumed common across origin types):

@@

\begin{equation}
\log\left(\hat{P}^{a}_{y,i=\mathrm{fall},j}\right) \sim \Ndist\left[\log\left(P^{a}_{y,i=\mathrm{fall},j} \right), \hat{\sigma}_{P^{a}_{y,i=\mathrm{fall},j}}\right]
(\#eq:fall-count-data)
\end{equation}

@@

\begin{equation}
\log\left(\hat{M}^{b}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}\right) \sim \Ndist\left[\log\left(M^{b}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}\right), \hat{\sigma}_{M^{b}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}}\right]
(\#eq:spring-count-data)
\end{equation}

@@

\begin{equation}
\logit\left(\hat{\phi}^{P^{b} \to M^{a}}_{y,j}\right) \sim \Ndist\left[\logit\left(\frac{\sum_{i}^{n_i} M^{a}_{y,i,o=\mathrm{NOR},j}}{P^{b}_{y,j}}\right),\hat{\sigma}_{\phi^{P^{b} \to M^{a}}_{y,j}}\right]
(\#eq:summer-surv-data)
\end{equation}

@@

\begin{equation}
\logit\left(\hat{\phi}^{P^{a} \to M^{a}}_{y,i=\mathrm{fall},j}\right) \sim \Ndist\left[\logit\left(\frac{M^{a}_{y,i=\mathrm{fall},o=\mathrm{NOR},j}}{P^{a}_{y,i=\mathrm{fall},j}}\right),\hat{\sigma}_{\phi^{P^{a} \to M^{a}}_{y,i=\mathrm{fall},j}}\right]
(\#eq:fall-surv-data)
\end{equation}

@@

\begin{equation}
\logit\left(\hat{\phi}^{P^{a} \to M^{a}}_{y,i=\mathrm{spring},j}\right) \sim \Ndist\left[\logit\left(\frac{M^{a}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}}{P^{a}_{y,i=\mathrm{spring},j} \cdot \phi^{P^{b} \to P^{a}}_{j}}\right),\hat{\sigma}_{\phi^{P^{a} \to M^{a}}_{y,i=\mathrm{spring},j}}\right]
(\#eq:winter-surv-data)
\end{equation}

for the winter tagging group (where $\phi^{P^{b} \to P^{a}}_j$ is a time-constant model-estimated survival from summer tagging to winter tagging for parr that ultimately become spring migrants and $j \in [\mathrm{CAT}, \mathrm{LOS}, \mathrm{UGR}]$),

\begin{equation}
\logit\left(\hat{\phi}^{M^{b} \to M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}\right) \sim \Ndist\left[\logit\left(\phi^{M^{b} \to M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}\right),\hat{\sigma}_{\phi^{M^{b} \to M^{a}}_{y,i=\mathrm{spring},o=\mathrm{NOR},j}}\right]
(\#eq:spring-surv-data)
\end{equation}

@@

\begin{equation}
\logit\left(\hat{\phi}^{M^{b} \to M^{a}}_{y,i=\mathrm{spring},o=\mathrm{HOR},j}\right) \sim \Ndist\left[\logit\left(\phi^{M^{b} \to M^{a}}_{y,i=\mathrm{spring},o=\mathrm{HOR},j}\right),\hat{\sigma}_{\phi^{M^{b} \to M^{a}}_{y,i=\mathrm{spring},o=\mathrm{HOR},j}}\right]
(\#eq:spring-surv-HOR-data)
\end{equation}

for hatchery-origin smolt releases ($j \in [\mathrm{CAT}, \mathrm{LOS}, \mathrm{UGR}]$).

Much quantitative research has been devoted to studying the survival of Chinook salmon smolt migrating through the hydrosystem located on the Snake and Columbia rivers and we thought it best to utilize the findings from some of this research as input data to our modeling effort rather than attempt to have our model freely estimate survival along this migration. Using external estimates of survival here so would give us the flexibility to more freely estimate quantities which have received less directed research, namely ocean survival.
We used the output from a recent Comparative Survival Study report [@mccann-etal-2020, Table A.1 on page A-20 therein] to serve as empirical observations (with estimates of uncertainty) with which to inform this component of our model.
Provided by @mccann-etal-2020 are annual estimates of in-stream survival along the migration from LGR through Bonneville Dam (BON); separate time series are available for natural-origin (we used the "Aggregate Wild Chinook" estimates as none were available for Grande Ronde populations only, denoted $\hat{\phi}^{M^{a} \to O^{0}}_{y,o=\mathrm{NOR}}$ here) and for hatchery-origin (we used the "Catherine Creek AP" estimates, denoted $\hat{\phi}^{M^{a} \to O^{0}}_{y,o=\mathrm{HOR}}$ here).
We treated these estimates as representative of the survival for each population, which implies an assumption that the different populations we model experience similar conditions during this migration and ignores any effect of barge-transporting fish.
Similar to the other survival data sets, we assumed these estimates are made with logit-normal random error around the life cycle model estimate to build the likelihood:

@@

We assumed first year natural-origin^[All $o=\mathrm{NOR}$ in eq. \@ref(eq:phi-O0-O1-random), omitted for brevity.] ocean survival was a multivariate logit-normal random variable around a time-constant expected value ($\dot{\phi}^{O^{0} \to O^{1}}_{o=\mathrm{NOR},j}$) with covariance matrix $\Sigma_{\phi^{O^{0} \to O^{1}}}$, but included a lag-1 autoregressive process as a means to account for ocean conditions that may affect survival in a temporally non-independent fashion\todo{Include reference to papers that talk about ocean survival varying cyclically; e.g., Mantua et al. paper on PDO}:

@@

We assumed process variation was negligible in the second and third year ocean survival terms, thus we forced the year-specific values of $\phi^{O^{1} \to O^{2}}_{y,o=\mathrm{NOR},j}$ and $\phi^{O^{2} \to O^{3}}_{y,o=\mathrm{NOR},j}$ to take on the same value for all years (i.e., $\dot{\phi}^{O^{1} \to O^{2}}_{o=\mathrm{NOR},j}$ and $\dot{\phi}^{O^{2} \to O^{3}}_{o=\mathrm{NOR},j}$, respectively).
Because not all maturity and ocean survival terms were uniquely identifiable based on adult return abundance and age composition, we relied on reasonably strong priors to inform $\dot{\phi}^{O^{1} \to O^{2}}_{o=\mathrm{NOR},j}$ and $\dot{\phi}^{O^{2} \to O^{3}}_{o=\mathrm{NOR},j}$ (Table \@ref(tab:param-table)).
\todo{Are we using priors or fixed values? Include reference to reports where we can reference the specific values used.}
To obtain hatchery-origin ocean survival each year, we adjusted the natural-origin survival by a log-odds ratio ($\delta_{j}$) that was common across all ocean years and brood years but varied by population:

\begin{equation}
\begin{aligned}
\logit\left(\phi^{O^{0} \to O^{1}}_{y,o=\mathrm{HOR},j}\right) &= \logit\left(\phi^{O^{0} \to O^{1}}_{y,o=\mathrm{NOR},j}\right) + \delta_j \\
\logit\left(\phi^{O^{1} \to O^{2}}_{y,o=\mathrm{HOR},j}\right) &= \logit\left(\phi^{O^{1} \to O^{2}}_{y,o=\mathrm{NOR},j}\right) + \delta_j \\
\logit\left(\phi^{O^{2} \to O^{3}}_{y,o=\mathrm{HOR},j}\right) &= \logit\left(\phi^{O^{2} \to O^{3}}_{y,o=\mathrm{NOR},j}\right) + \delta_j \\
\end{aligned}
(\#eq:get-phi-O-HOR)
\end{equation}

@@

Upon arrival to the natal tributary, populations with a weir and hatchery program (all populations except the Minam River, indexed by $j = \mathrm{MIN}$) have some number of adults removed each year for broodstock and as a means to minimize natural spawning of hatchery-origin fish.
We supplied these weir removals (denoted by $B_{y,k,o,j}$) to the model as known without error and set all $B_{y,k,o,j=\mathrm{MIN}} = 0$.
Similarly, fish are harvested in the natal tributaries prior to spawning, and we additionally supplied estimates of fish harvested ($H_{y,k,o,j}$).
Thus, the abundance of adults that would have a chance to spawn each year was:

@@

\begin{equation}
\zeta_{k,j=\mathrm{MIN}} = \mathrm{mean}(\zeta_{k,j \in [\mathrm{CAT}, \mathrm{LOS}, \mathrm{UGR}]})
(\#eq:get-zeta-minam)
\end{equation}

>>>>

---
title: "\\textsc{Mathematical Description}"
subtitle: "\\textsc{State-Space Life Cycle Model for Grande Ronde Spring Chinook Salmon}"
output: 
  bookdown::pdf_document2:
    keep_tex: true
    toc: true
    toc_depth: 3
    number_sections: true
    citation_package: natbib
  # bookdown::word_document2
fontfamily: mathpazo
geometry: [top=1in,bottom=1in,left=0.25in,right=1.75in]
linkcolor: cyan
urlcolor: cyan
citecolor: black
bibliography: cites.bib
link-citations: true
biblio-style: apalike
editor_options: 
  chunk_output_type: console
header-includes:
  - \usepackage{microtype}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \definecolor{my-blue}{HTML}{E6F0FF}
  - \usepackage[textwidth=1.5in]{todonotes}
  - \setuptodonotes{backgroundcolor=my-blue,textcolor=black,linecolor=ProcessBlue,bordercolor=my-blue,size=footnotesize}
  - \usepackage[labelfont={bf,sc},labelsep=period]{caption}
  - \captionsetup{width=\textwidth}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \cfoot{\thepage}
  - \renewcommand{\headrulewidth}{0.5pt}
  - \renewcommand{\footrulewidth}{0.5pt}
  - \lfoot{State-Space Life Cycle Model\\Mathematical Description}
  - \rfoot{Grande Ronde Spring Chinook Salmon}
  - \usepackage{titlesec}
  - \titleformat{\section}{\bfseries\LARGE\scshape}{\thesection}{1em}{}
  - \titleformat{\subsection}{\bfseries\large}{\thesubsection}{1em}{}
  - \titleformat{\subsubsection}{\itshape}{\thesubsubsection}{1em}{}
  - \renewcommand{\contentsname}{Table of Contents}
  - \renewcommand{\eqref}[1]{\ref{#1}}
  - \renewcommand*{\log}{\mathrm{log}\hspace{-0.20em}}
  - \renewcommand*{\max}{\mathrm{max}\hspace{-0.20em}}
  - \newcommand{\logit}{\mathrm{logit}\hspace{-0.20em}}
  - \newcommand{\ilogit}{\mathrm{logit}^{-1}\hspace{-0.20em}}
  - \newcommand{\MVNdist}{\mathcal{MVN}\hspace{-0.20em}}
  - \newcommand{\Ndist}{\mathcal{N}\hspace{-0.20em}}
  - \newcommand{\Bdist}{\mathcal{B}\hspace{-0.20em}}
  - \newcommand{\Mdist}{\mathcal{M}\hspace{-0.20em}}
  - \newcommand{\fall}{\mathrm{fall}}
  - \newcommand{\spring}{\mathrm{spring}}
  - \newcommand{\NOR}{\mathrm{NOR}}
  - \newcommand{\HOR}{\mathrm{HOR}}
  - \newcommand{\CAT}{\mathrm{CAT}}
  - \newcommand{\LOS}{\mathrm{LOS}}
  - \newcommand{\MIN}{\mathrm{MIN}}
  - \newcommand{\UGR}{\mathrm{UGR}}
  - \newcommand{\WUL}{\mathrm{WUL}}
---

@@

To facilitate later analyses (not documented here) investigating the effects of habitat restoration and climate change on population dynamics, we modeled parr capacity as a function of weighted usable habitat length ($\WUL_j$)^[Derivation of $\WUL_j$ is too detailed to describe fully here.
Briefly, the relative ability of stream segments with different habitat characteristics to hold rearing parr was expressed in terms of the density of fish predicted to reside there by a complex generalized linear mixed model.
The model was fitted to density estimates based on detectability-adjusted [@staton-etal-2022] summer snorkel counts with local habitat characteristics (e.g., pool frequency, water temperature, large wood density, etc.) as predictor variables.
The weight of each stream segment was assigned based on the predicted density relative value to the segment with the highest value; $\WUL_j$ was then equal to the sum of segment-specific weight and length within the known spawning and rearing extent for each population.].
We modeled parr recruitment capacity ($\beta_j$) as a function of this $\WUL_j$ metric:

\begin{equation}
\log\left(\beta_j\right) \sim \Ndist\left[\log\left(\lambda \cdot \WUL_j\right), \sigma_{\beta} \right]
(\#eq:beta-random)
\end{equation}

where $\lambda$ is the expected change in parr capacity per 1 km change in weighted usable habitat length, and $\sigma_{\beta}$ is the log-normal standard deviation of variability in this relationship not captured by $\WUL$ values.

@@

\begin{equation}
\log\left(\dot{L}^{P^{b}}_{y,j}\right) = \omega_{0,j} + \omega_{1,j} \cdot \log\left(\frac{E_{y,j}}{\WUL_j}\right)
(\#eq:get-L-Pb-dot)
\end{equation}

@@

Like many populations of stream-type spring Chinook salmon in the Columbia River basin \todo{Insert references about other CRB populations displaying the fall/spring migratory strategy}, there is life history diversity among the juveniles with respect to their migratory phenology and habitat types occupied for rearing overwinter.
Some portion of the summer parr ($P^{b}_{y,j}$)  migrate out of the headwaters near where they were spawned in the fall and rear farther downstream (termed "fall migrants" and indexed by $i = \fall$) and the remaining portion migrate out the following spring (termed "spring migrants" and indexed by $i = \spring$).
Due to this difference in migratory phenology, these two groups of fish are monitored separately \todo{Insert a reference to a recent ODFW screw trap monitoring report where it is stated (or implied) that migratory strategies are monitored separately.} depending on when they pass the rotary screw trap located in each tributary\todo{Create a map and include reference it here}.
To enable building an observation model that could accommodate this feature of the populations in our model, we apportioned summer parr to each of these two strategies:

@@

where $P^{a}_{y,i,j}$ is migratory strategy-specific parr abundance and $\pi_{y,i,j}$ is the proportion that take on each migratory strategy.
We assumed the expected value of this proportion for fall migrants ($\dot{\pi}_{i = \fall,j}$) was time-constant and modeled the realized values as having a multivariate logit-normal distribution with covariance matrix $\Sigma_{\pi_{i = \fall}}$:

\begin{equation}
\logit\left(\boldsymbol{\pi}_{y,i = \fall,1:n_j}\right) \sim \MVNdist\left[\logit\left(\boldsymbol{\dot{\pi}}_{i = \fall,1:n_j}\right), \Sigma_{\pi_{i = \fall}}\right]
(\#eq:pi-random)
\end{equation}

Since there are only two migratory strategies, we obtained the proportion that are spring migrants as the complement ($\pi_{y,i = \spring,j} = 1 - \pi_{y,i = \fall,j}$).

@@

Since hatchery-origin smolts are released into the population in spring (see Section \@ref(hatchery-inputs)), we introduce the origin index ($o$) starting at the in-basin smolt stage prior to out-migration ($M^{b}_{y,i,o,j}$).
The abundance of natural-origin (indexed by $o=\NOR$) smolt was then:

\begin{equation}
M^{b}_{y,i,o=\NOR,j} = P^{a}_{y,i,j} \cdot \phi^{P^{a} \to M^{b}}_{y,i,j}
(\#eq:get-Mb-NOR)
\end{equation}

$\rho_{\pi_{i=\fall,j,j^{\prime}}}$

@@

For populations with hatchery supplementation (CAT, LOS, and UGR), hatchery-origin smolt releases (assumed known without error \todo{Insert reference to a document that gives info about these smolt releases}) were stored in the variable $M^{b}_{y,i=\spring,o=\HOR,j}$; we assumed no hatchery-origin parr were released the prior year, thus $M^{b}_{y,i=\fall,o=\HOR,j} = 0$.
For all subsequent life stages, we tracked the abundance (and often survival) of hatchery-origin fish separately from natural-origin fish.

@@

All in-basin smolt ($M^{b}_{y,i,o,j}$) migrate to the ocean in the spring of the year they become age-2 regardless of their migratory strategy ($i$).
We divide mortality experienced during this migration into two stages: (_i_) from in-basin to Lower Granite Dam (LGR), the first in a series of eight dams\todo{Insert reference to a document describing dams} encountered in the downstream migration, and (_ii_) from LGR to the ocean.
For parameter identifiability in mortality stage (_i_), we were required to assume that survival from in-basin to LGR was identical between natural-origin fall and spring migrants (i.e., $\phi^{M^{b} \to M^{a}}_{y,i=\fall,o=\NOR,j} = \phi^{M^{b} \to M^{a}}_{y,i=\spring,o=\NOR,j}$, we thus omit the $i$ dimension when defining these quantities below), but we were able to model natural-origin survival separately from hatchery-origin because they are monitored separately (see Section \@ref(survival-data-sources)).

@@

\begin{equation}
\logit\left(\dot{\phi}^{M^{b} \to M^{a}}_{y,o=\NOR,j}\right) = \tau_{0,j} + \tau_{1,j} \cdot L^{*M^{b}}_{y,j}
(\#eq:get-phi-Mb-Ma-dot)
\end{equation}

where $L^{*M^{b}}_{y,j}$ has been scaled and centered by subtracting the historical average observed smolt mean length for population $j$ from $L^{M^{b}}_{y,j}$ and dividing by the historical observed standard deviation (just as for $L^{*P^{b}}_{y,j}$ for standardized parr mean length).
For hatchery-origin smolt, we assumed this expected out-of-basin survival term was time-constant (i.e., all $\dot{\phi}^{M^{b} \to M^{a}}_{y,o=\HOR,j}$ were equal within a population).
For each origin separately, we modeled realized out-of-basin migration survival as a multivariate logit-normal random variable around the expected value with covariance matrix $\Sigma_{\phi^{M^{b} \to M^{a}}}$ (assumed common across origin types):

@@

\begin{equation}
\log\left(\hat{P}^{a}_{y,i=\fall,j}\right) \sim \Ndist\left[\log\left(P^{a}_{y,i=\fall,j} \right), \hat{\sigma}_{P^{a}_{y,i=\fall,j}}\right]
(\#eq:fall-count-data)
\end{equation}

@@

\begin{equation}
\log\left(\hat{M}^{b}_{y,i=\spring,o=\NOR,j}\right) \sim \Ndist\left[\log\left(M^{b}_{y,i=\spring,o=\NOR,j}\right), \hat{\sigma}_{M^{b}_{y,i=\spring,o=\NOR,j}}\right]
(\#eq:spring-count-data)
\end{equation}

@@

\begin{equation}
\logit\left(\hat{\phi}^{P^{b} \to M^{a}}_{y,j}\right) \sim \Ndist\left[\logit\left(\frac{\sum_{i}^{n_i} M^{a}_{y,i,o=\NOR,j}}{P^{b}_{y,j}}\right),\hat{\sigma}_{\phi^{P^{b} \to M^{a}}_{y,j}}\right]
(\#eq:summer-surv-data)
\end{equation}

@@

\begin{equation}
\logit\left(\hat{\phi}^{P^{a} \to M^{a}}_{y,i=\fall,j}\right) \sim \Ndist\left[\logit\left(\frac{M^{a}_{y,i=\fall,o=\NOR,j}}{P^{a}_{y,i=\fall,j}}\right),\hat{\sigma}_{\phi^{P^{a} \to M^{a}}_{y,i=\fall,j}}\right]
(\#eq:fall-surv-data)
\end{equation}

@@

\begin{equation}
\logit\left(\hat{\phi}^{P^{a} \to M^{a}}_{y,i=\spring,j}\right) \sim \Ndist\left[\logit\left(\frac{M^{a}_{y,i=\spring,o=\NOR,j}}{P^{a}_{y,i=\spring,j} \cdot \phi^{P^{b} \to P^{a}}_{j}}\right),\hat{\sigma}_{\phi^{P^{a} \to M^{a}}_{y,i=\spring,j}}\right]
(\#eq:winter-surv-data)
\end{equation}

for the winter tagging group (where $\phi^{P^{b} \to P^{a}}_j$ is a time-constant model-estimated survival from summer tagging to winter tagging for parr that ultimately become spring migrants and $j \in [\CAT, \LOS, \UGR]$),

\begin{equation}
\logit\left(\hat{\phi}^{M^{b} \to M^{a}}_{y,i=\spring,o=\NOR,j}\right) \sim \Ndist\left[\logit\left(\phi^{M^{b} \to M^{a}}_{y,i=\spring,o=\NOR,j}\right),\hat{\sigma}_{\phi^{M^{b} \to M^{a}}_{y,i=\spring,o=\NOR,j}}\right]
(\#eq:spring-surv-data)
\end{equation}

@@

\begin{equation}
\logit\left(\hat{\phi}^{M^{b} \to M^{a}}_{y,i=\spring,o=\HOR,j}\right) \sim \Ndist\left[\logit\left(\phi^{M^{b} \to M^{a}}_{y,i=\spring,o=\HOR,j}\right),\hat{\sigma}_{\phi^{M^{b} \to M^{a}}_{y,i=\spring,o=\HOR,j}}\right]
(\#eq:spring-surv-HOR-data)
\end{equation}

for hatchery-origin smolt releases ($j \in [\CAT, \LOS, \UGR]$).

Much quantitative research has been devoted to studying the survival of Chinook salmon smolt migrating through the hydrosystem located on the Snake and Columbia rivers and we thought it best to utilize the findings from some of this research as input data to our modeling effort rather than attempt to have our model freely estimate survival along this migration. Using external estimates of survival here so would give us the flexibility to more freely estimate quantities which have received less directed research, namely ocean survival.
We used the output from a recent Comparative Survival Study report [@mccann-etal-2020, Table A.1 on page A-20 therein] to serve as empirical observations (with estimates of uncertainty) with which to inform this component of our model.
Provided by @mccann-etal-2020 are annual estimates of in-stream survival along the migration from LGR through Bonneville Dam (BON); separate time series are available for natural-origin (we used the "Aggregate Wild Chinook" estimates as none were available for Grande Ronde populations only, denoted $\hat{\phi}^{M^{a} \to O^{0}}_{y,o=\NOR}$ here) and for hatchery-origin (we used the "Catherine Creek AP" estimates, denoted $\hat{\phi}^{M^{a} \to O^{0}}_{y,o=\HOR}$ here).
We treated these estimates as representative of the survival for each population, which implies an assumption that the different populations we model experience similar conditions during this migration and ignores any effect of barge-transporting fish.
Similar to the other survival data sets, we assumed these estimates are made with logit-normal random error around the life cycle model estimate to build the likelihood:

@@

We assumed first year natural-origin^[All $o=\NOR$ in eq. \@ref(eq:phi-O0-O1-random), omitted for brevity.] ocean survival was a multivariate logit-normal random variable around a time-constant expected value ($\dot{\phi}^{O^{0} \to O^{1}}_{o=\NOR,j}$) with covariance matrix $\Sigma_{\phi^{O^{0} \to O^{1}}}$, but included a lag-1 autoregressive process as a means to account for ocean conditions that may affect survival in a temporally non-independent fashion\todo{Include reference to papers that talk about ocean survival varying cyclically; e.g., Mantua et al. paper on PDO}:

@@

We assumed process variation was negligible in the second and third year ocean survival terms, thus we forced the year-specific values of $\phi^{O^{1} \to O^{2}}_{y,o=\NOR,j}$ and $\phi^{O^{2} \to O^{3}}_{y,o=\NOR,j}$ to take on the same value for all years (i.e., $\dot{\phi}^{O^{1} \to O^{2}}_{o=\NOR,j}$ and $\dot{\phi}^{O^{2} \to O^{3}}_{o=\NOR,j}$, respectively).
Because not all maturity and ocean survival terms were uniquely identifiable based on adult return abundance and age composition, we relied on reasonably strong priors to inform $\dot{\phi}^{O^{1} \to O^{2}}_{o=\NOR,j}$ and $\dot{\phi}^{O^{2} \to O^{3}}_{o=\NOR,j}$ (Table \@ref(tab:param-table)).
\todo{Are we using priors or fixed values? Include reference to reports where we can reference the specific values used.}
To obtain hatchery-origin ocean survival each year, we adjusted the natural-origin survival by a log-odds ratio ($\delta_{j}$) that was common across all ocean years and brood years but varied by population:

\begin{equation}
\begin{aligned}
\logit\left(\phi^{O^{0} \to O^{1}}_{y,o=\HOR,j}\right) &= \logit\left(\phi^{O^{0} \to O^{1}}_{y,o=\NOR,j}\right) + \delta_j \\
\logit\left(\phi^{O^{1} \to O^{2}}_{y,o=\HOR,j}\right) &= \logit\left(\phi^{O^{1} \to O^{2}}_{y,o=\NOR,j}\right) + \delta_j \\
\logit\left(\phi^{O^{2} \to O^{3}}_{y,o=\HOR,j}\right) &= \logit\left(\phi^{O^{2} \to O^{3}}_{y,o=\NOR,j}\right) + \delta_j \\
\end{aligned}
(\#eq:get-phi-O-HOR)
\end{equation}

@@

Upon arrival to the natal tributary, populations with a weir and hatchery program (all populations except the Minam River, indexed by $j = \MIN$) have some number of adults removed each year for broodstock and as a means to minimize natural spawning of hatchery-origin fish.
We supplied these weir removals (denoted by $B_{y,k,o,j}$) to the model as known without error and set all $B_{y,k,o,j=\MIN} = 0$.
Similarly, fish are harvested in the natal tributaries prior to spawning, and we additionally supplied estimates of fish harvested ($H_{y,k,o,j}$).
Thus, the abundance of adults that would have a chance to spawn each year was:

@@

\begin{equation}
\zeta_{k,j=\MIN} = \mathrm{mean}(\zeta_{k,j \in [\CAT, \LOS, \UGR]})
(\#eq:get-zeta-minam)
\end{equation}
